Group polarization is often a mirage caused by inappropriate statistical analyses

### 2022-11-16 GroupPolarizationStatmod/papers/ms.tex
t-test table being made on small examples

When simulated data reliably result in false positive $t$-test results (where
the $t$-test $p \le 0.1$) this strengthens the case that

@@

\paragraph{\citet{Abrams1990}}

\paragraph{\citet{Burnstein1973}}

@@

>>>>

When simulated data reliably result in false positive $t$-test results (where
the $t$-test $p \le 0.1$) this strengthens the case that.

One complicating factor to this approach is that statistical power, $W$, varies
between treatments. Therefore, simulated data should yield statistically-significant
group polarization with frequency close to $W$ for a given treatment at a given
significance value, $\alpha$.

@@

\paragraph{\citet{Abrams1990}} This study is meant to support the presence of the "self-categorization" effect in
social psychology. The paper presents three experiments and their results--Experiment 3 is the group polarization
experiment, designed to show that self-categorization is at work in the emergence of group polarization. Uses high
schoolers aged 16-17 in Hawkes Bay, New Zealand. There are five deliberation topics (p. 112) across two conditions
(p. 113): one condition where participants are made aware of the relationship between their opinions and other group
members' opinions ("Categorized") and another condition where participants were given no information about their
opinion relative to other group members' opinions ("Uncategorized"). It was confusing to me what scale is used for
Table 2 (p. 114), but parsing the "Dependent measures" subsection several times led me to understand the last
paragraph, that "[a]ttitude shift overall was...assessed...on the unconverted +4 to -4 scale."

\paragraph{\citet{Burnstein1973}} First I want to note this study was funded in part by a Guggenheim fellowship (see
note on first page) and the authors acknolwedge the assistance of "vinous" discussions with colleagues at the
University of Provence and the "spirited and intelligent efforts" of two female research assistants. The study
presents the results of two experiments (really one experiment with two pools of participants) that test the
persuasive arguments hypothesis by having students perform the standard group polarization experiment with five CDQ
discussion items, items A,B,E,F, and H (A,B risk-inducing; E, F caution-inducing; H neutral). Experiment 1 used 149
male intro psych students and Experiment 2 used 76 male intro psych students from the University of Michigan. There
are three conditions across the two experiments--one hypothesized to show an emergent shift and the other two
hypothesized to not show a shift. None of the conditions allowed participants to know each others' ratings exactly,
the participants only deliberated the various topics. In the one hypothesized to show a shift, participants were
instructed to argue for their position. In the other two, participants argued against their position. The "against"
condition was further divided into a case where everyone knew everyone was prompted to argue against their position,
and another sub-condition where participants did not know if others were arguing for or against their true position.
We include only the first condition where participants argued for their position, which is the only condition
expected to show shifts (though others do, but the authors let themselves off the hook saying, "They are curious
findings and defy straightforward explanation...we will merely note their occurrence and not venture to speculate as
to their cause"). Thus, in our table we have conditions A, B, E, F, and H. Even though H is supposed to not show a
shift, it does in this study, which is possibly further evidence that this study merely shows participant consensus
formation, not opinion shifts.

\paragraph{\citet{Burnstein1975}} In this study, Burnstein and Vinokur ran an experiment to evaluate their theory
that generating persuasive arguments are necessary for group polarization to occur. The authors design a study with
three conditions, one Experimental and two Control, where participants (60 male students from intro psych at the
University of Michigan) first answered four CDQ items that typically generate shifts towards more risk, then, in the
Experimental condition, saw group members' responses and produced arguments that might support the positions of
other participants. In the first control condition, Condition II in the paper and Control1 in our table,
participants were exposed to others' opinions, but does not get to reflect on their opinions or give arguments. In
condition III, the second control condtition (we call Control2) participants do not get to know other participants'
opinions at all. Thus, Burnstein and Vinokur hypothesize that significant shifts will only occur in Condition I, the
Experimental condition. We thus only include Condition I in our analysis. The authors only report the shift, meaning
we are free to choose any pre- and post-deliberation value to see if the null hypothesis is actually plausible.

\paragraph{\citet{Schkade2010}} Friedkin uses three CDQ topics, "Sports", "School", and "Surgery" that use a
20-point CDQ risk scale in fractional units, 0.05, 0.10, ..., 1.00 across three group sizes (2, 3, and 4, which he
calls "dyads", "triads", and "tetrads", which we use in our CaseStudyTag). Friedkin only gives opinion shifts, and
apparently transforms the fractional CDQ responses into percentages, since the shifts range between 5 and 9.
However, closer inspection of Friedkin's results shows that the reported "Choice Shift" is not equal to the
difference between mean final and initial opinions. Friedkin also uses "different" means for pre- and
post-discussion opinions, which seems easy enough to show are really the same by re-organizing terms in the sum.
Friedkin uses two topics in the tetrad condition where responses are in units of money where participants can freely
give any number they choose ("Asbestos" and "Disaster"), and another topic called "Institutions" where participants
rate 13 institutions, which participants rated on a scale from 0 to 100 percent to indicate the degree of trust the
participant has in each institution. We only consider the CDQ topics since these are the ones rated on a discrete
scale.

@@

\paragraph{\citet{Krizan2007}} Uses five CDQ topics ("scenarios") from Kogan and
Wallach (1967) that typically provide shifts toward more risky decisions.  These are
given in Footnote 2 on p. 194. Krizan and Baron report shifts and mean pre-test
opinions on the risk scale, meaning that shifts are expected to be negative towards
lower odds of success required to take the risk, i.e., the acceptance of more risk.
Krizan and Baron report mean pre-test opinions, but average over all three conditions
for each scenario, then report shifts for each scenario and condition, so we
effectively have no way for an exact match of initial mean to condition x scenario,
where conditions are no-discussion, deliberation without knowledge of an out-group's
opinion, and deliberation with knowledge of an out-group's opinion. We are only
interested in the non-control cases--K\&B found no difference in knowing out-group
opinion, which they say is evidence against the self-categorization hypothesis of group
polarization.  Therefore, we have ten cases we test across five scenarios and two
conditions with our model, and we use the reported initial mean for each scenario
twice--once for each scenario across the two conditions. Post-deliberation opinions are
calculated in the spreadsheet using the reported shifts. K\&B also report the "Overall"
shift, but I find this meaningless since it is across different items with no
statistical control for inter-item variability. I noted that false detections are
plausible in all cases due to the fact that the standard deviations of shifts overwhelm
the shifts themselves, but I will still run the model on these cases to confirm.

\paragraph{\citet{Moscovici1969}} Moscovici and Zavalloni studied whether group
polarization is more general than CDQ by using opinions about political topics. There
were two general topics--attitudes about "Americans" and attitudes about then-president
of France Charles "DeGaulle"--with 12 and 11 specific items, respectively, on which
participants gave opinions and deliberated. Participants deliberated on one of the two
general topics, giving opinions on each item on a seven-point Likert scale, -3 to +3.
The authors reported pre-consensus and post-consensus mean opinions taken over all
participants and items for each of the two general topics I have called the "Americans"
and "DeGaulle" conditions.

\paragraph{\citet{Myers1970}} Myers and Bishop test whether group polarization occurs
in groups with different levels of racial bias determined by a survey that is not also
the deliberation topic. The participants were high school students from western
Michigan. In the experimental condition, social influence occurs through deliberation;
in the control condition there is no deliberation or social influence at all, just
re-tested opinions after a distraction task. Obviously we only report on whether there
was a significant shift in the experimental condition. In both conditions, three groups
were created according to their racial prejudice as measured by a 100-item racial
attitude measurement instrument, called "low", "medium", and "high" prejudice groups.
The participants then gave their opinions, deliberated, and gave opinions again on
eight deliberation topics. Means were taken across items and prejudice group.

\paragraph{\citet{Myers1975}} Myers performs two experiments each with experimental and control conditions for two group types with initial biases. In Experiment 1, Myers creates biased groups by having participants (intro psychology students) read a description of six different teachers, three "bad" professors and three "good" professors, which the participants are assumed to infer through vignettes about the professors. The participants in the experimental condition of Experiment 1 deliberated in small groups after giving their initial opinions on six items a 0-to-10 bad-to-good scale. In Experiment 2, participants were put into conservatively or liberally biased small groups based on their pre-deliberation opinions on six deliberation items regarding womens' rights and the role of women in the family and society. Again, in the experimental group, participants deliberated each item before giving their final opinions on the items. In the control group in each Experiment, there was no deliberation, just some filler task. Curiously, significant shifts were still found in the control groups--it is unclear to us why. Experiment 1 also had participants adjust the bad and good professors' pay to obtain non-ordinal measures of opinion regarding these professors, which we do not consider here. However, as we will discuss in the paper, this does not make it free of problems--like so many other group polarization studies, Myers statistical model does not account for inter-item, inter-group, or inter-participant variability.

\paragraph{\citet{Schkade2010}} Participants in Schkade, Sunstein, and Hastie's study were put into two politically ideologically biased groups based on their geographical location in the US state of Colorado: participants were drawn from Boulder, Colorado (abbreviated CO), to be liberally biased and from Colorado Springs, CO, to be conservatively biased. Participants gave their opinions and discussed three topics related to Global Warming, Civil Unions, and Affirmative Action. Pre- and post-deliberation mean are given in Table 1 (p. 232) for all six (topics x geographies).
